---
title: "Reproducible Research: Peer Assessment 1"
author: "Johan Vásquez Mazo"
output: 
  html_document:
    keep_md: true
---

First off, global options are set so that every code chunk is echoed.

```{r setoptions, echo = TRUE}
library(knitr)
opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data

The *activity.zip* file only contains a file: *activity.csv*, so the data are easily read by using the function `read.csv` and unzipping the file from within the function. The activity monitoring data are stored into a data frame called `data`.

```{r loaddata}
data <- read.csv(file = unz("activity.zip", "activity.csv"), header = TRUE)
str(data)
```

By calling `str(data)`, it is seen that there are three variables: `steps`, `date` and `interval`; and 17568 observations. However, the `date` variable is of class `character`, so it needs to be set as a date variable by using the `as.Date` function, with argument `format` set equal to `"%Y-%m-%d"`.

```{r dateclass}
data$date <- as.Date(data$date, format = "%Y-%m-%d")
str(data)
```

The data are now tidy and ready to be processed.

## What is mean total number of steps taken per day?

To calculate the total number of steps taken per day, the function `tapply` comes in handy by using the `date` variable as a factor. A list called `totalSteps` is created when calling `tapply` with arguments `X = data$Steps`, `INDEX = data$date`, `FUN = sum` and `na.rm = TRUE`. Since `tapply` coerces the `INDEX` elements into factors, it is not necessary to call `as.factor(data$date)` when setting the `INDEX` argument. Afterwards, this list `totalSteps` is converted into a data frame by using the `as.data.frame` function.

```{r totalsteps}
totalSteps <- tapply(data$steps, data$date, sum, na.rm = TRUE)
totalSteps <- as.data.frame(totalSteps)
summary(totalSteps)
```

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
